\documentclass[12pt]{article}

\usepackage{sbc-template}

\usepackage{graphicx,url}

\usepackage[T1]{fontenc}
\usepackage[brazilian,portuges]{babel}
\usepackage[utf8]{inputenc}

\sloppy

\title{A Systematic Mapping Study on Quality of Service in Service-Oriented Computing}
\author{Danilo Filgueira Mendon\c{c}a\inst{1}, Gena\'{i}na Nunes Rodrigues\inst{1}, \\ Rodrigo Bonif\'{a}cio\inst{1}, Alet\'{e}ia Favacho\inst{1}, Maristela Holanda\inst{1} }

\address{Department of Computer Science -- University of Brasilia (UnB) -- Campus Darcy Ribeiro\\
70910-900, Brasilia, DF, Brazil
\email{dfmendonca@gmail.com, \{genaina, rbonifacio, aleteia, mholanda\}@cic.unb.br}
}

\begin{document} 

\maketitle

\begin{abstract}
  In the last years, the field of service oriented computing (SOC) has received a growing interest from researchers and practitioners, particularly with respect to quality of service (QoS). For instance, searching for terms regarding QoS in the context of SOC in major digital libraries (e.g. IEEExplore and ACM Digital) we find more than 700 contributions only in the last four years. In such a massive field, a mapping study (MS) is paramount to identify trends and research opportunities, guide research efforts and avoid redundant contributions. This paper presents a MS to aggregate literature in this field. Our major findings show that, with respect to SOC contributions dealing with QoS properties, most of them concentrate on monitoring and adaptation, while very few focus on coordination and communication. Regarding quality attributes, a vast majority the papers use generic models, so that the proposed solutions are independent of the particularities of a quality attribute. Nevertheless, we find that availability and performance were the major highlights, differently from cost models, scalability and security. With respect to research type, most of the reviewed studies propose new solutions, instead of evaluating and validating existing proposals--- a symptom of a field that does not follow established research paradigms.
\end{abstract}

\begin{resumo}
  Nos últimos anos, o campo da computação orientada a serviços (SOC) recebeu um crescente interesse de pesquisadores e praticantes, particularmente no que diz respeito a qualidade de serviços (QoS). Por exemplo, a busca por termos como QoS e qualidade de serviços no contexto de SOC em bibliotecas consagradas (e.g. EEExplore e ACM Digital), são encontradas mais de 700 publicações somente nos últimos quatro anos. Numa área tão extensa, um mapeamento de estudos é crucial para identificar tendências e oportunidades de pesquisas, guiar esforços de pesquisa e evitar contribuições redundantes. Esse artigo apresenta um MS para agrupar a literatura neste campo. Nossas maiores descobertas mostram que, em relação às contribuições para SOC envolvendo propriedades de QoS, a maioria se concentra em monitoramento e adaptação, enquanto poucas focam em coordenação e comunicação. A respeito dos atributos de qualidade, uma vasta maioria dos artigos utilizam modelos genéricos, de modo que suas contribuições são independentes das particularidades de um dado atributo. Não obstante, descobrimos que disponibilidade e desempenho foram os principais destaques, ao contrário de custo, escalabilidade e segurança. A respeito do tipo de pesquisa, a maioria dos estudos avalizados propuseram novas soluções em vez de avaliarem ou validarem propostas existentes --- um sintoma de um campo que não segue paradigmas de pesquisa estabelecidos.
  
\end{resumo}

\input{dados}

\input{introduction}
    
\input{review_method}

\input{results}

\input{ameacas}

\input{conclusion}


\bibliographystyle{sbc}
\bibliography{sbc-template}

\end{document}
